---
title: "DESeq"
author: "Zaneta Ku"
date: "2023-11-14"
output: pdf_document
---

# 1. Bioconductor and DESeq2 setup

```{r}
library(BiocManager)
library(DESeq2)
```

# 2. Import countData and colData

read the count data 

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
head(counts)
head(metadata)
```

## Q1. How many genes are in this dataset? 

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names = 1)
number_of_genes <- nrow(counts)
print(number_of_genes)

```

## Q2. How many ‘control’ cell lines do we have? 

```{r}
metadata <- read.csv("airway_metadata.csv")
number_of_control_samples <- sum(metadata$condition_column == "control")
print(number_of_control_samples)

```


# 3. Toy differential gene expression

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

```{r}
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```


## Q3. How would you make the above code in either approach more robust?

```{r}
library(dplyr)

control <- metadata %>% filter(dex == "control")
if (nrow(control) == 0) {
  stop("No control samples found or incorrect column name for 'dex'")
}
control.counts <- counts %>% select(all_of(control$id))
control.mean <- rowSums(control.counts, na.rm = TRUE) / nrow(control)
print(head(control.mean))



```

## Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r}
library(dplyr)

treated <- metadata %>% filter(dex == "treated")
if (nrow(treated) == 0) {
  stop("No treated samples found or incorrect column name for 'dex'")
}
treated.counts <- counts %>% select(all_of(treated$id))
treated.mean <- rowSums(treated.counts, na.rm = TRUE) / nrow(treated)
print(head(treated.mean))

```

```{r}
meancounts <- data.frame(control.mean, treated.mean)

```

## Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

```{r}
plot(meancounts[,1], meancounts[,2], xlab="Control", ylab="Treated")

```

## Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot?

```{r}
library(ggplot2)
ggplot(meancounts, aes(x=control.mean, y=treated.mean)) + 
    geom_point() + 
    labs(x="Control", y="Treated")

```
I would use geom_point() to create a scatter plot.

## Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this? 

```{r}
plot(meancounts[,1], meancounts[,2], xlab="Control", ylab="Treated", log="xy")

```

```{r}
ggplot(meancounts, aes(x=control.mean, y=treated.mean)) + 
    geom_point() +
    scale_x_continuous(trans="log2") +
    scale_y_continuous(trans="log2") +
    labs(x="Control", y="Treated")

```
Calculating Log2 Fold Change and Handling Zero Values

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
zero.vals <- which(meancounts[,1:2] == 0, arr.ind = TRUE)
to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)

```

## Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

The arr.ind=TRUE argument causes which() to return both row and column indices where the condition is TRUE. In this case, it helps identify genes (rows) and samples (columns) with zero counts. Using unique() on the first column of the output ensures that each gene is only counted once, even if it has zero counts in multiple samples.

## Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 

```{r}
up.ind <- mycounts$log2fc > 2
sum(up.ind)  # Counts the number of TRUE elements in up.ind

```

## Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level? 

```{r}
down.ind <- mycounts$log2fc < (-2)
sum(down.ind)  # Counts the number of TRUE elements in down.ind

```

## Q10. Do you trust these results? Why or why not?

The results should be approached with caution. While we have identified genes with significant fold changes, this analysis does not account for statistical significance (like p-values). Therefore, the results might be misleading.


# 4. DESeq2 analysis

Import data

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

DESeq analysis

```{r}
dds <- DESeq(dds)
```

results

```{r}
res <- results(dds)
res
```
summarize

```{r}
summary(res)
```

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

# 5. Adding annotation data

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
```

```{r}
head(res)
```

## Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res$entrez, res$uniprot and res$genename.

```{r}
# Add Entrez ID
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), 
                     keytype="ENSEMBL",
                     column="ENTREZID",
                     multiVals="first")

# Add UniProt Accession
res$uniprot <- mapIds(org.Hs.eg.db,
                      keys=row.names(res),
                      keytype="ENSEMBL",
                      column="UNIPROT",
                      multiVals="first")

# Add Gene Name
res$genename <- mapIds(org.Hs.eg.db,
                       keys=row.names(res),
                       keytype="ENSEMBL",
                       column="GENENAME",
                       multiVals="first")
```

```{r}
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```

```{r}
write.csv(res[ord,], "deseq_results.csv")
```

# 6. Data Visualization

volcano plots 

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

add guideline

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

color vector

```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)

```

more customization

```{r}
library(EnhancedVolcano)
```

```{r}
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```

# 7. Pathway analysis

load package

```{r}
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)

# Examine the first 2 pathways in this kegg set for humans
head(kegg.sets.hs, 2)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

run analysis 

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```

```{r}
# Look at the first three down (less) pathways
head(keggres$less, 3)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa05310")

```

```{r}
# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa05310", kegg.native=FALSE)
```

## Q12. Can you do the same procedure as above to plot the pathview figures for the top 2 down-reguled pathways?

Identifying Top 2 Down-Regulated Pathways

```{r}
top_down_pathways <- head(keggres$less, 2)
top_down_pathway_ids <- rownames(top_down_pathways)

print("Top 2 Down-Regulated Pathway IDs:")
print(top_down_pathway_ids)

print("Details of the Top 2 Down-Regulated Pathways:")
print(top_down_pathways)
```

Plotting Pathview Figures

```{r}

library(pathview)

# Plot for the first down-regulated pathway
pathview(gene.data=foldchanges, pathway.id=top_down_pathway_ids[1])

# Plot for the second down-regulated pathway
pathview(gene.data=foldchanges, pathway.id=top_down_pathway_ids[2])

print(paste("Plotting pathview for pathway:", top_down_pathway_ids[2]))
pathview(gene.data=foldchanges, pathway.id=top_down_pathway_ids[2])
```

