---
title: "RNASeq"
author: "Zaneta Ku"
date: "2023-11-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

```{r}
# Read the data file using read.table()
data <- read.table("bioboot.github.io_bggn213_W19_class-material_rs8067378_ENSG00000172057.6.txt", 
                   header = TRUE, sep = " ")

# Display a summary of the data
summary(data)

# Calculating sample size for each genotype
sample_size <- table(data$geno)

# Generating a boxplot and saving the output to an object
bp <- boxplot(exp ~ geno, data = data, plot = FALSE)

# Median expression levels for each genotype can be found in the bp object
median_expression <- bp$stats[3,]

# Display the results
print("Sample sizes for each genotype:")
print(sample_size)
print("Median expression levels for each genotype:")
print(median_expression)

```

# Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
# Generating the boxplot
boxplot(exp ~ geno, data = data, main = "Expression by Genotype",
        xlab = "Genotype", ylab = "Expression Level", 
        col = c("lightblue", "lightgreen", "pink"))

```
The boxplot suggests that the A/A genotype is associated with higher ORMDL3 expression levels compared to the G/G genotype, indicating that the SNP may influence gene expression. The A/G genotype shows intermediate expression levels, supporting a potential additive genetic effect on ORMDL3 expression.
